Normalization by Evaluation for \\ Intuitionistic Propositional Logic

### 2018-07-17 notes/nbeSum.tex
ITC: evaluation

The interpretation $\den{A \imp B}$ of implication will follow the
Brouwer-Heyting-Kolmogorov (BHK) interpretation of intuitionistic
logic:
\begin{center}\itshape
  A proof of $A \imp B$ is a method turning any proof of $A$ into a
  proof of $B$.
\end{center}
However, the direct lifting of the functions space $\den{A \imp
  B}_\Gamma = \den A_\Gamma \to \den B_\Gamma$ does not work.  It does
not give a presheaf because of the contravariant occurrence of
presheaf $\den A$.  Instead, we force monotonicity by quantifying over
all OPEs of $\Gamma$:
\[
  \den{A \imp B}_\Gamma = \prod_{\tau : \Delta \leq \Gamma} (\den A
  _\Delta \to \den B _\Delta)
\]
Reification is also challenging.  Let $f : \den{A \imp B}_\Gamma$ and
attempt
%$\tau : \Delta \leq \Gamma$
\[
  \reify[\Gamma]{A \imp B} f %\;\Delta\;\tau =
    = \impIP {\reify[\Gamma.A]B{(f\;(\weak[A]{\id[\Gamma]})}\;\mbox{?})}
\]
where hole ``?'' should be filled with an element of $\den B_{\Gamma.A}$
which intuitively should represent the new hypothesis $\top :
\Hyp[\Gamma.A]A$ introduced by $\timpI$.  Thus we need a natural
transformation $\Hyp A \todot \den A$ which \emph{reflects} variables
into the semantics.  In order to define it by
induction on $A$, we need to generalize it to
\[
  \treflect^A : \Ne A \todot \den A
  .
\]
Let us make sure we can define it at atoms and conjunctions:
\[
\begin{array}{lllll}
  \treflect^P & = & \tne & : & \Ne P \todot \Nf P \\
  \treflect^\top & = & !  & : & \Ne \top \todot \hatone \\
  \treflect^{A \land B} & = & [\, \treflect^A \comp \andE1  , \;
                             \treflect^B \comp \andE2 ]
    & : & \Ne{A \land B} \todot \den A \hattimes \den B \\
\end{array}
\]
Observe the eliminations $\andE i$ introduced to reflect conjunctive hypotheses!

@@

>>>>

The interpretation $\den{A \imp B}$ of implication will follow the
Brouwer-Heyting-Kolmogorov (BHK) interpretation of intuitionistic
logic:
\begin{center}\itshape
  A proof of $A \imp B$ is a method turning any proof of $A$ into a
  proof of $B$.
\end{center}
However, the direct lifting of the functions space $\den{A \imp
  B}_\Gamma = \den A_\Gamma \to \den B_\Gamma$ does not work.  It does
not give a presheaf because of the contravariant occurrence of
presheaf $\den A$.  Instead, we force monotonicity by quantifying over
all OPEs of $\Gamma$:
\[
  \den{A \imp B}_\Gamma = \prod_{\Delta : \Cxt} \prod_{\tau : \Delta \leq \Gamma} (\den A
  _\Delta \to \den B _\Delta)
\]
Reification is also challenging.  Let $f : \den{A \imp B}_\Gamma$ and
attempt
%$\tau : \Delta \leq \Gamma$
\[
  \reify[\Gamma]{A \imp B} f %\;\Delta\;\tau =
    = \impIP {\reify[\Gamma.A]B{(f\;(\weak[A]{\id[\Gamma]})}\;\mbox{?})}
\]
where hole ``?'' should be filled with an element of $\den B_{\Gamma.A}$
which intuitively should represent the new hypothesis $\top :
\Hyp[\Gamma.A]A$ introduced by $\timpI$.  Thus we need a natural
transformation $\Hyp A \todot \den A$ which \emph{reflects} variables
into the semantics.  In order to define it by
induction on $A$, we need to generalize it to
\[
  \treflect^A : \Ne A \todot \den A
  .
\]
Let us make sure we can define it at atoms and conjunctions:
\[
\begin{array}{lllll}
  \treflect^P & = & \tne & : & \Ne P \todot \Nf P \\
  \treflect^\top & = & !  & : & \Ne \top \todot \hatone \\
  \treflect^{A \land B} & = & [\, \treflect^A \comp \andE1  , \;
                             \treflect^B \comp \andE2 ]
    & : & \Ne{A \land B} \todot \den A \hattimes \den B \\
\end{array}
\]
Observe the eliminations $\andE i$ introduced to reflect conjunctive hypotheses!

@@

Evaluation is turning an expression $t : \TmG A$ into a value in
$\den A$.  However, a direct definition $\teval^A : \Tm A \todot \den A$
by induction on the expression fails in case $\timpI$:
\[
\begin{array}{lll}
   \teval^{A \imp B}_\Gamma & : & \TmG A \to \den {A \imp B} _\Gamma \\
   \teval^{A \imp B}\,(\impI t)\;\Delta\;a & = & ? \\
\end{array}
\]
To be able to pass the argument $a$ to the function $\teval\,(\impI
t)$, we have to generalize evaluation to take an valuation of the
context into account.
\[
  \teval^{A \imp B}_\Gamma  : \den \Gamma \todot \den {A \imp B}
\]
Here, $\den \Gamma$ is the presheaf obtained as the product of the
hypotheses contained in $\Gamma$:
\[
\begin{array}{lll}
  \den{\cempty} & = & \hatone \\
  \den{\Gamma.A} & = & \den\Gamma \hattimes \den A \\
\end{array}
\]
Looking up values in the environment is the following family of
natural transformations:
\[
\begin{array}{l@{~}lcl}
  \multicolumn 2 l {\tlookup^A_\Gamma} & : & \HypG A \to (\den \Gamma \todot \den A) \\
  \lookup[\Gamma.A]A \ttop    & \Delta\,(\gamma,a) & = & a \\
  \lookup[\Gamma.B]A {\pop x} &\Delta\,(\gamma,a) & = & \lookup[\Gamma]A x\, \Delta\, \gamma \\
\end{array}
\]
\begin{exercise}
  Express $\tlookup$ via the projections
  $\pi_i : \A_1 \hattimes \A_2 \todot \A_i$
  out of the presheaf product.
\end{exercise}
We are ready to define evaluation:
\[
\begin{array}{lll}
  \evalG A {\hyp x} & = & \lookupG A x \\
  \evalG {A \imp B} {\impI t}\,\Delta\,\gamma\,\Delta'\,(\tau : \Delta' \leq \Delta)\,a
    & = & \eval[\Gamma.A]B t\,\Delta'\,(\den\Gamma_\tau\gamma,a) \\
  \evalG {B} {\impE t u}\,\Delta\,\gamma & = &
    \evalG {A \imp B} t\,\Delta\,\gamma\,\Delta\,\id[\Delta]\,
      (\evalG A u\,\Delta\,\gamma) \\
\end{array}
\]
\begin{exercise}
  Complete the definition of $\teval$ for conjunctions.
\end{exercise}
\begin{exercise}
  Express $\teval$ in terms of the operations of the CCC of
  presheaves!
\end{exercise}

\subsection{Digression: The CCC of reifyable presheaves}

Instead of defining the semantics $\den A$ of formula $A$ by induction
on $A$, we can consider a category whose objects are interpretations
of formulas, and show that this category is cartesian closed.

An object in this category would be a quadruple consisting of:
\[
\begin{array}{lll@{\qquad}l}
  \A & : & \PSh  & \mbox{a presheaf} \\
  A  & : & \Form & \mbox{a formula} \\
  \treflect^A & : & \Ne A \todot \A
    & \mbox{a method to reflect neutrals into $\A$} \\
  \treify^A & : & \A \todot \Nf A
    & \mbox{a method to reify values of $\A$} \\
\end{array}
\]
\begin{exercise}
  Complete this definition by a suitable notion of morphism.
\end{exercise}
\begin{exercise}
  Show that this category is cartesian closed.
\end{exercise}

\section{Extension to the positive connectives}
